---
title: "Activity 07: XGBoost Quantile Forecast"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

Activity 07 tests an XGBoost quantile-regression alternative to the original
ARIMA forecast. It keeps the same testing season, reference dates, horizons,
target, location, and FluSight quantile ladder, but replaces `auto.arima()` with
direct horizon-specific gradient-boosted trees.

The model intentionally uses no wastewater, NSSP, or other leading indicators.
Its inputs are lagged influenza admissions and deterministic target-week
seasonal harmonics only. This makes it a clean machine-learning benchmark
between the baseline ARIMA work and the later ARIMAX/blended models.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.width = 10, fig.height = 6)

find_project_dir <- function(start_dir = getwd()) {
  candidates <- unique(c(
    start_dir,
    file.path(start_dir, "USE-ME", "carolmoreira"),
    file.path(start_dir, ".."),
    file.path(start_dir, "..", ".."),
    file.path(start_dir, "..", "..", "..")
  ))
  candidates <- normalizePath(candidates, mustWork = FALSE)
  candidates <- candidates[file.exists(file.path(candidates, "output/scripts/07_xgboost_forecast.R"))]
  if (!length(candidates)) stop("Could not locate USE-ME/carolmoreira.")
  candidates[1]
}

project_dir <- find_project_dir()
knitr::opts_knit$set(root.dir = project_dir)
if (!isTRUE(getOption("knitr.in.progress"))) setwd(project_dir)
cat("Working directory:", project_dir, "\n")
```

## Run The Workflow

The script below rebuilds the forecast, scores it, compares it with available
benchmarks, and writes the figures used in this report.

```{r run-script}
source("output/scripts/07_xgboost_forecast.R")
```

## Load Outputs

```{r packages}
required <- c("readr", "dplyr", "tidyr", "ggplot2", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Required R package(s) not installed: ", paste(missing, collapse = ", "))

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

```{r paths}
forecast_csv <- "output/data/07_xgboost/xgboost_flusight_forecasts.csv"
scores_csv <- "output/data/07_xgboost/xgboost_forecast_scores.csv"
summary_csv <- "output/data/07_xgboost/xgboost_summary_by_horizon.csv"
comparison_csv <- "output/data/07_xgboost/xgboost_model_comparison_by_horizon.csv"
spec_csv <- "output/data/07_xgboost/xgboost_specification.csv"
diagnostics_csv <- "output/data/07_xgboost/xgboost_diagnostics.csv"

forecast_png <- "output/figures/07_xgboost/xgboost_forecast_vs_observed.png"
comparison_png <- "output/figures/07_xgboost/xgboost_metric_comparison.png"
width_png <- "output/figures/07_xgboost/xgboost_interval_widths.png"
feature_png <- "output/figures/07_xgboost/xgboost_feature_overview.png"

expected <- c(forecast_csv, scores_csv, summary_csv, comparison_csv,
              spec_csv, diagnostics_csv, forecast_png, comparison_png,
              width_png, feature_png)
if (!all(file.exists(expected))) {
  stop("Missing Activity 07 output(s): ", paste(expected[!file.exists(expected)], collapse = ", "))
}
```

## Model Design

XGBoost has no built-in autoregressive time-series structure, so the weekly
series is converted into a supervised learning table. For each reference date
and horizon, the model uses the last four known admissions values plus first
and second seasonal harmonics for the target MMWR week.

```{r model-spec}
read_csv(spec_csv, show_col_types = FALSE) %>%
  kable(caption = "XGBoost quantile model specification.")
```

```{r feature-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(feature_png)))
```

## No-Look-Ahead Framing

The XGBoost model is fit directly by horizon. For a reference date `r` and
horizon `h`, training examples use only anchor weeks whose observed target
`t + 7h` is no later than `r`. Prediction uses the anchor week `r`, so all lag
features are known at forecast time.

```{r diagnostics}
diagnostics <- read_csv(diagnostics_csv, show_col_types = FALSE)
diagnostics %>%
  group_by(horizon) %>%
  summarise(
    reference_dates = n(),
    min_training_rows = min(n_training_rows),
    max_training_rows = max(n_training_rows),
    .groups = "drop"
  ) %>%
  kable(caption = "Training row counts by direct forecast horizon.")
```

## FluSight Forecast Output

Each reference date emits 69 rows: 3 horizons times 23 quantiles. Quantiles are
clamped at zero, rounded to integer counts, sorted to prevent crossing, and
validated against the FluSight date and ladder rules.

```{r forecast-preview}
forecasts <- read_csv(forecast_csv, show_col_types = FALSE)
forecasts %>%
  arrange(reference_date, horizon, output_type_id) %>%
  head(12) %>%
  kable(caption = "Preview of the Activity 07 FluSight-format XGBoost forecast.")
```

```{r forecast-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(forecast_png)))
```

## XGBoost Performance

```{r summary-table}
xgb_summary <- read_csv(summary_csv, show_col_types = FALSE)
xgb_summary %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  kable(caption = "XGBoost forecast performance by horizon.")
```

```{r width-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(width_png)))
```

## Comparison With Other Versions

The comparison table includes the original ARIMA baseline, Activity 5 ensemble,
and Activity 8 final blend when those outputs are available locally. XGBoost is
not the best one-week model, but it performs strongly at two and three weeks
ahead in this run.

```{r comparison-table}
comparison <- read_csv(comparison_csv, show_col_types = FALSE)
comparison %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  arrange(horizon, mean_wis) %>%
  kable(caption = "Activity 07 XGBoost compared with available forecast versions.")
```

```{r comparison-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(comparison_png)))
```

## Interpretation

XGBoost behaves differently from ARIMA-style models. It can learn nonlinear
relationships between recent admissions, season timing, and future admissions,
which helps at longer horizons here. The tradeoff is that it is less
transparent, it does not guarantee horizon-wise interval widening, and its
short-horizon forecast is less competitive than the blend and Activity 5
ensemble.

```{r strengths-weaknesses}
tibble::tribble(
  ~aspect, ~assessment,
  "Strength",
  "Strong two- and three-week WIS in this evaluation window.",
  "Strength",
  "Uses the exact same FluSight format and reference dates as the baseline.",
  "Strength",
  "Direct quantile regression allows asymmetric predictive distributions.",
  "Weakness",
  "No external leading indicators, so it cannot use wastewater or NSSP signal.",
  "Weakness",
  "Less interpretable than ARIMA or ARIMAX.",
  "Weakness",
  "Direct horizon models do not always produce intervals that widen monotonically."
) %>%
  kable(caption = "Summary of Activity 07 strengths and weaknesses.")
```

## Reproducibility Summary

```{r reproducibility-summary}
cat("Forecast CSV:", forecast_csv, "\n")
cat("Scores CSV:", scores_csv, "\n")
cat("Comparison CSV:", comparison_csv, "\n")
cat("Rows in forecast:", nrow(forecasts), "\n")
cat("Reference dates:", dplyr::n_distinct(forecasts$reference_date), "\n")
cat("Rows per reference date:", nrow(forecasts) / dplyr::n_distinct(forecasts$reference_date), "\n")
```
