---
title: "Activity 07: XGBoost Quantile Forecast"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

Activity 07 tests XGBoost quantile-regression alternatives to the original
ARIMA forecast. It keeps the same testing season, reference dates, horizons,
target, location, and FluSight quantile ladder, but replaces `auto.arima()` with
direct horizon-specific gradient-boosted trees.

Two XGBoost candidates are compared:

- Admissions-only XGBoost: lagged influenza admissions plus target-week
  seasonal harmonics.
- XGBoost + leading indicators: the same features plus NWSS WVAL and NSSP
  influenza ED visit signals lagged six weeks from the target week.

The leading indicators are lagged so that every feature is available before the
forecast reference date.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.width = 10, fig.height = 6)

find_project_dir <- function(start_dir = getwd()) {
  candidates <- unique(c(
    start_dir,
    file.path(start_dir, "USE-ME", "carolmoreira"),
    file.path(start_dir, ".."),
    file.path(start_dir, "..", ".."),
    file.path(start_dir, "..", "..", "..")
  ))
  candidates <- normalizePath(candidates, mustWork = FALSE)
  candidates <- candidates[file.exists(file.path(candidates, "output/scripts/07_xgboost_forecast.R"))]
  if (!length(candidates)) stop("Could not locate USE-ME/carolmoreira.")
  candidates[1]
}

project_dir <- find_project_dir()
knitr::opts_knit$set(root.dir = project_dir)
if (!isTRUE(getOption("knitr.in.progress"))) setwd(project_dir)
cat("Working directory:", project_dir, "\n")
```

## Run The Workflow

The script below rebuilds the forecast, scores it, compares it with available
benchmarks, and writes the figures used in this report.

```{r run-script}
source("output/scripts/07_xgboost_forecast.R")
```

## Load Outputs

```{r packages}
required <- c("readr", "dplyr", "tidyr", "ggplot2", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Required R package(s) not installed: ", paste(missing, collapse = ", "))

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

```{r paths}
forecast_csv <- "output/data/07_xgboost/xgboost_flusight_forecasts.csv"
leading_forecast_csv <- "output/data/07_xgboost/xgboost_leading_flusight_forecasts.csv"
scores_csv <- "output/data/07_xgboost/xgboost_forecast_scores.csv"
leading_scores_csv <- "output/data/07_xgboost/xgboost_leading_forecast_scores.csv"
summary_csv <- "output/data/07_xgboost/xgboost_summary_by_horizon.csv"
leading_summary_csv <- "output/data/07_xgboost/xgboost_leading_summary_by_horizon.csv"
comparison_csv <- "output/data/07_xgboost/xgboost_model_comparison_by_horizon.csv"
spec_csv <- "output/data/07_xgboost/xgboost_specification.csv"
diagnostics_csv <- "output/data/07_xgboost/xgboost_diagnostics.csv"
leading_diagnostics_csv <- "output/data/07_xgboost/xgboost_leading_diagnostics.csv"
tuning_csv <- "output/data/07_xgboost/xgboost_hyperparameter_tuning.csv"

forecast_png <- "output/figures/07_xgboost/xgboost_forecast_vs_observed.png"
leading_forecast_png <- "output/figures/07_xgboost/xgboost_leading_forecast_vs_observed.png"
comparison_png <- "output/figures/07_xgboost/xgboost_metric_comparison.png"
width_png <- "output/figures/07_xgboost/xgboost_interval_widths.png"
leading_width_png <- "output/figures/07_xgboost/xgboost_leading_interval_widths.png"
feature_png <- "output/figures/07_xgboost/xgboost_feature_overview.png"
leading_feature_png <- "output/figures/07_xgboost/xgboost_leading_feature_overview.png"
tuning_png <- "output/figures/07_xgboost/xgboost_hyperparameter_tuning.png"

expected <- c(forecast_csv, scores_csv, summary_csv, comparison_csv,
              spec_csv, diagnostics_csv, leading_forecast_csv,
              leading_scores_csv, leading_summary_csv, leading_diagnostics_csv,
              forecast_png, leading_forecast_png, comparison_png,
              width_png, leading_width_png, feature_png, leading_feature_png,
              tuning_csv, tuning_png)
if (!all(file.exists(expected))) {
  stop("Missing Activity 07 output(s): ", paste(expected[!file.exists(expected)], collapse = ", "))
}
```

## Model Design

XGBoost has no built-in autoregressive time-series structure, so the weekly
series is converted into a supervised learning table. For each reference date
and horizon, both candidates use the last four known admissions values plus
first and second seasonal harmonics for the target MMWR week. The leading
candidate additionally uses WVAL and NSSP values from six weeks before the
target week.

Hyperparameters are selected with a small rolling-validation grid using only
pre-test reference dates. This avoids choosing XGBoost settings using the
2025-26 evaluation outcomes.

```{r model-spec}
read_csv(spec_csv, show_col_types = FALSE) %>%
  kable(caption = "XGBoost quantile model specification.")
```

```{r tuning-table}
read_csv(tuning_csv, show_col_types = FALSE) %>%
  mutate(across(c(mean_wis, mae), round, 3)) %>%
  arrange(model, mean_wis) %>%
  kable(caption = "Pre-test rolling-validation hyperparameter results.")
```

```{r tuning-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(tuning_png)))
```

```{r feature-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(feature_png)))
```

```{r leading-feature-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(leading_feature_png)))
```

## No-Look-Ahead Framing

The XGBoost model is fit directly by horizon. For a reference date `r` and
horizon `h`, training examples use only anchor weeks whose observed target
`t + 7h` is no later than `r`. Prediction uses the anchor week `r`, so all lag
features are known at forecast time.

```{r diagnostics}
diagnostics <- read_csv(diagnostics_csv, show_col_types = FALSE)
leading_diagnostics <- read_csv(leading_diagnostics_csv, show_col_types = FALSE)
bind_rows(diagnostics, leading_diagnostics) %>%
  group_by(horizon) %>%
  summarise(
    models = n_distinct(model),
    reference_dates = n_distinct(reference_date),
    min_training_rows = min(n_training_rows),
    max_training_rows = max(n_training_rows),
    .groups = "drop"
  ) %>%
  kable(caption = "Training row counts by direct forecast horizon.")
```

## FluSight Forecast Output

Each reference date emits 69 rows: 3 horizons times 23 quantiles. Quantiles are
clamped at zero, rounded to integer counts, sorted to prevent crossing, and
validated against the FluSight date and ladder rules.

```{r forecast-preview}
forecasts <- read_csv(forecast_csv, show_col_types = FALSE)
forecasts %>%
  arrange(reference_date, horizon, output_type_id) %>%
  head(12) %>%
  kable(caption = "Preview of the Activity 07 FluSight-format XGBoost forecast.")
```

```{r leading-forecast-preview}
leading_forecasts <- read_csv(leading_forecast_csv, show_col_types = FALSE)
leading_forecasts %>%
  arrange(reference_date, horizon, output_type_id) %>%
  head(12) %>%
  kable(caption = "Preview of the Activity 07 XGBoost + leading indicators forecast.")
```

```{r forecast-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(forecast_png)))
```

```{r leading-forecast-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(leading_forecast_png)))
```

## XGBoost Performance

```{r summary-table}
xgb_summary <- read_csv(summary_csv, show_col_types = FALSE)
xgb_leading_summary <- read_csv(leading_summary_csv, show_col_types = FALSE)
bind_rows(xgb_summary, xgb_leading_summary) %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  arrange(horizon, mean_wis) %>%
  kable(caption = "XGBoost forecast performance by horizon.")
```

```{r width-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(width_png)))
```

```{r leading-width-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(leading_width_png)))
```

## Comparison With Other Versions

The comparison table includes both XGBoost candidates plus the original ARIMA
baseline, Activity 5 ensemble, and Activity 8 final forecast when those outputs
are available locally. In this run, adding leading indicators improved coverage
but widened intervals enough to worsen WIS relative to admissions-only XGBoost.

```{r comparison-table}
comparison <- read_csv(comparison_csv, show_col_types = FALSE)
comparison %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  arrange(horizon, mean_wis) %>%
  kable(caption = "Activity 07 XGBoost compared with available forecast versions.")
```

```{r comparison-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(comparison_png)))
```

## Interpretation

XGBoost behaves differently from ARIMA-style models. It can learn nonlinear
relationships between recent admissions, season timing, and future admissions,
which helps at longer horizons here. Adding WVAL/NSSP did not improve WIS in
this backtest, likely because the surveillance streams have a shorter aligned
history and produced much wider intervals. This is still useful evidence: the
leading signal is available, but it needs more careful regularization or a
targeted ensemble before it should replace admissions-only XGBoost.

After tuning, the selected settings are more regularized than the original
fixed default: shallower trees with column and row subsampling. That improves
the rigor of the comparison even when the final winner remains a horizon-level
ensemble in Activity 08.

```{r strengths-weaknesses}
tibble::tribble(
  ~aspect, ~assessment,
  "Strength",
  "Strong two- and three-week WIS in this evaluation window.",
  "Strength",
  "Uses the exact same FluSight format and reference dates as the baseline.",
  "Strength",
  "Direct quantile regression allows asymmetric predictive distributions.",
  "Strength",
  "A leakage-safe WVAL/NSSP XGBoost variant is now tested directly.",
  "Weakness",
  "The leading-indicator variant widened intervals and worsened WIS in this run.",
  "Weakness",
  "Less interpretable than ARIMA or ARIMAX.",
  "Weakness",
  "Direct horizon models do not always produce intervals that widen monotonically."
) %>%
  kable(caption = "Summary of Activity 07 strengths and weaknesses.")
```

## Reproducibility Summary

```{r reproducibility-summary}
cat("Forecast CSV:", forecast_csv, "\n")
cat("Leading forecast CSV:", leading_forecast_csv, "\n")
cat("Scores CSV:", scores_csv, "\n")
cat("Leading scores CSV:", leading_scores_csv, "\n")
cat("Comparison CSV:", comparison_csv, "\n")
cat("Rows in forecast:", nrow(forecasts), "\n")
cat("Rows in leading forecast:", nrow(leading_forecasts), "\n")
cat("Reference dates:", dplyr::n_distinct(forecasts$reference_date), "\n")
cat("Rows per reference date:", nrow(forecasts) / dplyr::n_distinct(forecasts$reference_date), "\n")
```
