---
title: "Activity 8: FluSight-Compliant Final Forecast"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

This activity turns the best-performing forecast from the expanded modeling
work into a FluSight-ready product. The workflow is deliberately conservative:
it preserves the Activity 3 reference dates, horizons, target definition, and
quantile format; uses only lagged surveillance signals; ranks candidate models
with held-out weighted interval score (WIS); and validates the final submission
files before they are written.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.width = 10, fig.height = 6)

find_project_dir <- function(start_dir = getwd()) {
  candidates <- unique(c(
    start_dir,
    file.path(start_dir, "USE-ME", "carolmoreira"),
    file.path(start_dir, ".."),
    file.path(start_dir, "..", ".."),
    file.path(start_dir, "..", "..", "..")
  ))
  candidates <- normalizePath(candidates, mustWork = FALSE)
  candidates <- candidates[file.exists(file.path(candidates, "output/scripts/08_flusight.R"))]
  if (!length(candidates)) stop("Could not locate USE-ME/carolmoreira.")
  candidates[1]
}

project_dir <- find_project_dir()
knitr::opts_knit$set(root.dir = project_dir)
if (!isTRUE(getOption("knitr.in.progress"))) setwd(project_dir)
cat("Working directory:", project_dir, "\n")
```

## Run The Workflow

The script below is the reproducible source of truth for Activity 8. It rebuilds
the ARIMAX candidates, blends them with the Activity 5 ensemble and original
ARIMA forecast, selects the best model by held-out mean WIS, and writes the
final FluSight files.

```{r run-script}
source("output/scripts/08_flusight.R")
```

## Load Outputs

```{r packages}
required <- c("readr", "dplyr", "tidyr", "ggplot2", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Required R package(s) not installed: ", paste(missing, collapse = ", "))

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

```{r paths}
truth_csv <- "output/data/01_cleaning/cleaned_flu_admissions.csv"
inventory_csv <- "output/data/08_flusight/upstream_data_stream_inventory.csv"
spec_csv <- "output/data/08_flusight/arimax_specification.csv"
comparison_csv <- "output/data/08_flusight/arimax_model_comparison_by_horizon.csv"
ranking_csv <- "output/data/08_flusight/flusight_model_ranking.csv"
final_forecast_csv <- "output/data/08_flusight/flusight_final_forecasts.csv"
submission_manifest_csv <- "output/data/08_flusight/final_flusight_submission_manifest.csv"
compliance_audit_csv <- "output/data/08_flusight/flusight_compliance_audit.csv"

stream_png <- "output/figures/08_flusight/surveillance_streams_overview.png"
lag_png <- "output/figures/08_flusight/lagged_covariate_relationships.png"
component_png <- "output/figures/08_flusight/blend_vs_component_medians.png"
ranking_png <- "output/figures/08_flusight/flusight_model_ranking.png"
final_forecast_png <- "output/figures/08_flusight/flusight_final_forecast_vs_observed.png"
metric_png <- "output/figures/08_flusight/arimax_metric_comparison.png"
calibration_png <- "output/figures/08_flusight/rolling_interval_calibration_multipliers.png"

expected <- c(truth_csv, inventory_csv, spec_csv, comparison_csv, ranking_csv,
              final_forecast_csv, submission_manifest_csv, compliance_audit_csv,
              stream_png, lag_png, component_png, ranking_png, final_forecast_png,
              metric_png, calibration_png)
if (!all(file.exists(expected))) {
  stop("Missing Activity 8 output(s): ", paste(expected[!file.exists(expected)], collapse = ", "))
}
```

## Inputs And Leading Signals

The final candidate set uses the cleaned national influenza hospitalization
target plus two upstream national surveillance streams: NWSS wastewater viral
activity level and NSSP smoothed influenza emergency-department visit percent.
Both external signals are lagged six weeks before modeling, which makes them
available before every 1-, 2-, and 3-week-ahead target date.

```{r inventory}
read_csv(inventory_csv, show_col_types = FALSE) %>%
  kable(caption = "Input stream inventory and model role.")
```

```{r stream-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(stream_png)))
```

```{r lag-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(lag_png)))
```

## Model Specification

The central ARIMAX candidate uses lagged NWSS WVAL, lagged NSSP influenza ED
visits, deterministic MMWR timing features, lagged current-season cumulative
admissions, and a lagged post-peak-decline flag. The post-peak feature is
computed from the peak observed so far, then lagged, so it does not use the
eventual full-season peak.

```{r specification}
read_csv(spec_csv, show_col_types = FALSE) %>%
  kable(caption = "ARIMAX specification used inside the final candidate set.")
```

## Candidate Ranking

The selected production forecast is the lowest mean WIS candidate across all
scored reference dates and horizons. WIS is the primary ranking metric because
it rewards both sharpness and calibration. MAE and RMSE are shown as supporting
central-forecast diagnostics, while coverage and interval width describe
uncertainty behavior.

```{r ranking-table}
ranking <- read_csv(ranking_csv, show_col_types = FALSE)
best_model <- ranking$model[1]
ranking %>%
  mutate(across(c(mean_wis, mae, rmse, coverage_95, mean_width_95), round, 3)) %>%
  kable(caption = "Overall model ranking. Lower WIS is better.")
```

```{r ranking-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(ranking_png)))
```

The final model selected by held-out mean WIS is:

```{r best-model}
cat(best_model)
```

## Horizon-Level Performance

```{r comparison-table}
comparison <- read_csv(comparison_csv, show_col_types = FALSE)
comparison %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  arrange(horizon, mean_wis) %>%
  kable(caption = "Model comparison by forecast horizon.")
```

```{r metric-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(metric_png)))
```

```{r component-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(component_png)))
```

## Interval Calibration

Rolling interval calibration was evaluated as an explicit candidate rather
than assumed to be better. It improved coverage for some horizons but increased
interval width substantially, so the final selection step allowed WIS to decide
whether that tradeoff was worthwhile.

```{r calibration-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(calibration_png)))
```

## Final Forecast

The final forecast keeps the required FluSight long format: one row per
reference date, target, horizon, location, output type, quantile level, and
forecast value.

```{r final-forecast-preview}
final_fc <- read_csv(final_forecast_csv, show_col_types = FALSE)
final_fc %>%
  arrange(reference_date, horizon, output_type_id) %>%
  head(12) %>%
  kable(caption = "Preview of the final FluSight forecast file.")
```

```{r final-forecast-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(final_forecast_png)))
```

## FluSight Compliance Audit

```{r compliance}
read_csv(compliance_audit_csv, show_col_types = FALSE) %>%
  kable(caption = "Automated compliance checks applied before final export.")
```

```{r submission-manifest}
manifest <- read_csv(submission_manifest_csv, show_col_types = FALSE)
manifest %>%
  kable(caption = "Per-reference-date FluSight submission files.")
```

The output directory contains one validated file per reference date. Each file
has 69 rows: 3 horizons times 23 quantiles.

## Strengths And Weaknesses

```{r strengths-weaknesses}
model_notes <- tibble::tribble(
  ~model, ~strengths, ~weaknesses,
  "Original ARIMA",
  "Simple, transparent, fully compliant, and hard to overfit.",
  "Uses no wastewater, syndromic, or seasonal state signal; reacts after admissions move.",
  "Activity 5 ensemble",
  "Stable central forecasts and narrower intervals than ARIMAX-style models.",
  "Does not directly use leading surveillance streams.",
  "ARIMAX: surveillance + seasonal",
  "Uses lagged upstream indicators and interpretable seasonal features.",
  "Central errors improved less consistently than the blend; intervals can be wide.",
  "Calibrated ARIMAX",
  "Uses only prior forecast errors to adjust uncertainty and improves coverage.",
  "Wider intervals can raise WIS when added coverage is not enough to compensate.",
  "Blend: 50% ARIMAX, 25% ensemble, 25% ARIMA",
  "Best overall WIS in this run; balances leading indicators with stable baseline behavior.",
  "Still inherits some missed coverage from component intervals.",
  "Calibrated blend",
  "Most conservative blend-style uncertainty and better coverage.",
  "Too wide for the WIS objective in this evaluation window."
)

model_notes %>%
  kable(caption = "Qualitative interpretation of each model version.")
```

## Reproducibility Summary

```{r reproducibility-summary}
cat("Final forecast CSV:", final_forecast_csv, "\n")
cat("Submission manifest:", submission_manifest_csv, "\n")
cat("Submission directory: output/data/08_flusight/final_flusight_submission\n")
cat("Selected model:", best_model, "\n")
cat("Reference dates:", dplyr::n_distinct(final_fc$reference_date), "\n")
cat("Rows in final forecast:", nrow(final_fc), "\n")
```
