---
title: "Activity 8: FluSight-Compliant Final Forecast"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

This activity turns the best-performing forecast from the expanded modeling
work into a FluSight-ready product. The workflow is deliberately conservative:
it preserves the Activity 3 reference dates, horizons, target definition, and
quantile format; compares ARIMA, Activity 5, Activity 7 XGBoost, Activity 7
XGBoost with leading indicators, ARIMAX, calibrated, blended,
horizon-specific, phase-calibrated, and reconciled candidates with held-out
weighted interval score (WIS); and validates the final submission files before
they are written.

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = TRUE, fig.width = 10, fig.height = 6)

find_project_dir <- function(start_dir = getwd()) {
  candidates <- unique(c(
    start_dir,
    file.path(start_dir, "USE-ME", "carolmoreira"),
    file.path(start_dir, ".."),
    file.path(start_dir, "..", ".."),
    file.path(start_dir, "..", "..", "..")
  ))
  candidates <- normalizePath(candidates, mustWork = FALSE)
  candidates <- candidates[file.exists(file.path(candidates, "output/scripts/08_flusight.R"))]
  if (!length(candidates)) stop("Could not locate USE-ME/carolmoreira.")
  candidates[1]
}

project_dir <- find_project_dir()
knitr::opts_knit$set(root.dir = project_dir)
if (!isTRUE(getOption("knitr.in.progress"))) setwd(project_dir)
cat("Working directory:", project_dir, "\n")
```

## Run The Workflow

The script below is the reproducible source of truth for Activity 8. It rebuilds
the ARIMAX candidates, blends them with the Activity 5 ensemble and original
ARIMA forecast, selects the best model by held-out mean WIS, and writes the
final FluSight files.

```{r run-script}
source("output/scripts/08_flusight.R")
```

## Load Outputs

```{r packages}
required <- c("readr", "dplyr", "tidyr", "ggplot2", "knitr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) stop("Required R package(s) not installed: ", paste(missing, collapse = ", "))

library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

```{r paths}
truth_csv <- "output/data/01_cleaning/cleaned_flu_admissions.csv"
inventory_csv <- "output/data/08_flusight/upstream_data_stream_inventory.csv"
spec_csv <- "output/data/08_flusight/arimax_specification.csv"
comparison_csv <- "output/data/08_flusight/arimax_model_comparison_by_horizon.csv"
ranking_csv <- "output/data/08_flusight/flusight_model_ranking.csv"
final_forecast_csv <- "output/data/08_flusight/flusight_final_forecasts.csv"
submission_manifest_csv <- "output/data/08_flusight/final_flusight_submission_manifest.csv"
compliance_audit_csv <- "output/data/08_flusight/flusight_compliance_audit.csv"
reconciliation_audit_csv <- "output/data/08_flusight/flusight_reconciliation_audit.csv"

stream_png <- "output/figures/08_flusight/surveillance_streams_overview.png"
lag_png <- "output/figures/08_flusight/lagged_covariate_relationships.png"
component_png <- "output/figures/08_flusight/blend_vs_component_medians.png"
ranking_png <- "output/figures/08_flusight/flusight_model_ranking.png"
final_forecast_png <- "output/figures/08_flusight/flusight_final_forecast_vs_observed.png"
metric_png <- "output/figures/08_flusight/arimax_metric_comparison.png"
calibration_png <- "output/figures/08_flusight/rolling_interval_calibration_multipliers.png"

expected <- c(truth_csv, inventory_csv, spec_csv, comparison_csv, ranking_csv,
              final_forecast_csv, submission_manifest_csv, compliance_audit_csv,
              reconciliation_audit_csv,
              stream_png, lag_png, component_png, ranking_png, final_forecast_png,
              metric_png, calibration_png)
if (!all(file.exists(expected))) {
  stop("Missing Activity 8 output(s): ", paste(expected[!file.exists(expected)], collapse = ", "))
}
```

## Inputs And Leading Signals

The final candidate set uses the cleaned national influenza hospitalization
target, the Activity 7 XGBoost forecasts, and two upstream national surveillance
streams: NWSS wastewater viral activity level and NSSP smoothed influenza
emergency-department visit percent. The external surveillance signals are
lagged six weeks before ARIMAX and XGBoost-leading modeling, which makes them
available before every 1-, 2-, and 3-week-ahead target date.

```{r inventory}
read_csv(inventory_csv, show_col_types = FALSE) %>%
  kable(caption = "Input stream inventory and model role.")
```

```{r stream-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(stream_png)))
```

```{r lag-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(lag_png)))
```

## Model Specification

The central ARIMAX candidate uses lagged NWSS WVAL, lagged NSSP influenza ED
visits, deterministic MMWR timing features, lagged current-season cumulative
admissions, and a lagged post-peak-decline flag. The Activity 7 XGBoost
candidate uses admissions lags and seasonal harmonics, and the XGBoost-leading
candidate adds lagged WVAL/NSSP. The post-peak ARIMAX feature is computed from
the peak observed so far, then lagged, so it does not use the eventual
full-season peak.

```{r specification}
read_csv(spec_csv, show_col_types = FALSE) %>%
  kable(caption = "ARIMAX specification used inside the final candidate set.")
```

## Candidate Ranking

The selected production forecast is the lowest mean WIS candidate across all
scored reference dates and horizons. WIS is the primary ranking metric because
it rewards both sharpness and calibration. The candidate set now includes a
horizon-specific model that uses the XGBoost-weighted blend for horizon 1 and
raw Activity 7 XGBoost for horizons 2 and 3, a low-weight raw/leading XGBoost
ensemble, and a phase-aware interval calibration candidate. MAE and RMSE are
shown as supporting central-forecast diagnostics, while coverage and interval
width describe uncertainty behavior.

```{r ranking-table}
ranking <- read_csv(ranking_csv, show_col_types = FALSE)
best_model <- ranking$model[1]
ranking %>%
  mutate(across(c(mean_wis, mae, rmse, coverage_95, mean_width_95), round, 3)) %>%
  kable(caption = "Overall model ranking. Lower WIS is better.")
```

```{r ranking-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(ranking_png)))
```

The final model selected by held-out mean WIS is:

```{r best-model}
cat(best_model)
```

## Horizon-Level Performance

```{r comparison-table}
comparison <- read_csv(comparison_csv, show_col_types = FALSE)
comparison %>%
  mutate(across(c(mae, rmse, mean_wis, coverage_95, mean_width_95), round, 3)) %>%
  arrange(horizon, mean_wis) %>%
  kable(caption = "Model comparison by forecast horizon.")
```

```{r metric-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(metric_png)))
```

```{r component-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(component_png)))
```

## Interval Calibration

Rolling interval calibration was evaluated as an explicit candidate rather
than assumed to be better. It improved coverage for some horizons but increased
interval width substantially. After Activity 7 was added, XGBoost-style direct
quantile forecasts were also allowed into the final ranking; for those models,
horizon-wise interval widening is reported diagnostically rather than treated
as a hard stop because direct per-horizon quantile models are not guaranteed to
widen monotonically.

The final refinement also adds phase-aware calibration. It estimates interval
multipliers from prior realized errors with the same horizon and epidemic phase
when enough historical examples are available, falling back to horizon-only
history when the phase-specific sample is too small.

```{r calibration-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(calibration_png)))
```

## Final Forecast

The final forecast keeps the required FluSight long format: one row per
reference date, target, horizon, location, output type, quantile level, and
forecast value.

```{r final-forecast-preview}
final_fc <- read_csv(final_forecast_csv, show_col_types = FALSE)
final_fc %>%
  arrange(reference_date, horizon, output_type_id) %>%
  head(12) %>%
  kable(caption = "Preview of the final FluSight forecast file.")
```

```{r final-forecast-plot, echo=FALSE, results='asis'}
cat(sprintf('<img src="%s" width="100%%" />', normalizePath(final_forecast_png)))
```

## FluSight Compliance Audit

```{r compliance}
read_csv(compliance_audit_csv, show_col_types = FALSE) %>%
  kable(caption = "Automated compliance checks applied before final export.")
```

```{r reconciliation}
read_csv(reconciliation_audit_csv, show_col_types = FALSE) %>%
  kable(caption = "Forecast reconciliation diagnostics for the selected final model.")
```

```{r submission-manifest}
manifest <- read_csv(submission_manifest_csv, show_col_types = FALSE)
manifest %>%
  kable(caption = "Per-reference-date FluSight submission files.")
```

The output directory contains one validated file per reference date. Each file
has 69 rows: 3 horizons times 23 quantiles.

## Strengths And Weaknesses

```{r strengths-weaknesses}
model_notes <- tibble::tribble(
  ~model, ~strengths, ~weaknesses,
  "Original ARIMA",
  "Simple, transparent, fully compliant, and hard to overfit.",
  "Uses no wastewater, syndromic, or seasonal state signal; reacts after admissions move.",
  "Activity 5 ensemble",
  "Stable central forecasts and narrower intervals than ARIMAX-style models.",
  "Does not directly use leading surveillance streams.",
  "Activity 7 XGBoost",
  "Best overall WIS after Activity 7 is included; especially strong at two and three weeks ahead.",
  "Less interpretable than ARIMA/ARIMAX and direct horizon intervals do not always widen monotonically.",
  "Activity 7 XGBoost + leading indicators",
  "Uses the requested WVAL/NSSP signals in a leakage-safe direct quantile model.",
  "Improved coverage but widened intervals and worsened WIS in this evaluation window.",
  "ARIMAX: surveillance + seasonal",
  "Uses lagged upstream indicators and interpretable seasonal features.",
  "Central errors improved less consistently than the blend; intervals can be wide.",
  "Calibrated ARIMAX",
  "Uses only prior forecast errors to adjust uncertainty and improves coverage.",
  "Wider intervals can raise WIS when added coverage is not enough to compensate.",
  "Blend: 50% ARIMAX, 25% ensemble, 25% ARIMA",
  "Balances leading indicators with stable baseline behavior.",
  "Still inherits some missed coverage from component intervals.",
  "Blend: 50% XGBoost, 25% ARIMAX, 15% ensemble, 10% ARIMA",
  "Second-best overall WIS after adding Activity 7; strong coverage and best one-week WIS in this run.",
  "Did not beat raw XGBoost overall because the added components weakened longer-horizon sharpness.",
  "Blend: 85% XGBoost, 15% XGBoost-leading",
  "Uses leading indicators conservatively rather than giving them equal weight.",
  "Still did not beat the horizon-specific final model by WIS.",
  "Horizon-specific: h1 XGBoost blend, h2-h3 XGBoost",
  "Uses the best observed one-week behavior from the XGBoost-weighted blend and the best longer-horizon behavior from Activity 7 XGBoost.",
  "More operationally complex because the production rule switches model source by horizon.",
  "Phase-calibrated horizon-specific",
  "Applies interval calibration by horizon and epidemic phase using prior errors only.",
  "Reduced interval width but did not improve WIS enough to beat the uncalibrated horizon-specific model.",
  "Calibrated blend",
  "Most conservative blend-style uncertainty and better coverage.",
  "Too wide for the WIS objective in this evaluation window."
)

model_notes %>%
  kable(caption = "Qualitative interpretation of each model version.")
```

## Reproducibility Summary

```{r reproducibility-summary}
cat("Final forecast CSV:", final_forecast_csv, "\n")
cat("Submission manifest:", submission_manifest_csv, "\n")
cat("Submission directory: output/data/08_flusight/final_flusight_submission\n")
cat("Selected model:", best_model, "\n")
cat("Reference dates:", dplyr::n_distinct(final_fc$reference_date), "\n")
cat("Rows in final forecast:", nrow(final_fc), "\n")
```
